---
title: "Non_NormlaisedNMR_ITU_Study - Final Plot Outputs for Sig Metabs Part 4"
author: "DJW & IJG"
date: "07/05/2020"
output: word_document
editor_options: 
  chunk_output_type: console
---

load libraries and packages
```{r}
library(tidyverse)
library(ggplot2)
library(ggfortify)
library(ggpubr)
library(pheatmap)
library(limma)
library(splines)
library(sjPlot)
```

## Create overall volcano plot of all basal comparisons for DE

```{r}
# Apply multiple testing correction and obtain stats
list_fitMetabs <- list(fitBasalTime_AA, fitBasalTime_Lipid, fitBasalTime_SmallMol)
volcanoPlot_list <- list() # to hold generated plots
plot_id <- c("AA", "Lipid", "Small Molecule")
labs <- c("Basal D10 vs D1", "Basal D10 vs D7", "Basal D7 vs D1")

for (i in 1:length(list_fitMetabs)) {
  stats_df_AA <- topTable(list_fitMetabs[[i]], number = Inf ) %>%
  tibble::rownames_to_column("Metabolite")

  # Let's extract the contrast p values for each and transform them with -log10()
  # contrast_p_vals_df_AA <- -log10(list_fitMetabs[[1]]$p.value) %>%
  contrast_p_vals_df_AA <- as.data.frame(list_fitMetabs[[i]]$p.value) %>%  # extract p values
    mutate(BasalD10_D1 = p.adjust(BasalD10_D1, method = "BH"),
           BasalD7_D1 = p.adjust(BasalD7_D1, method = "BH"), 
           BasalD10_D7 = p.adjust(BasalD10_D7, method = "BH")) # correct p values
    contrast_p_vals_df_AA <- -log10(contrast_p_vals_df_AA) %>%
    # Store metabolites as their own column
    tibble::rownames_to_column("Metabolite") %>%
    # Make this into long format
    tidyr::pivot_longer(dplyr::contains("Basal"),
                      names_to = "contrast",
                      values_to = "neg_log10_p_val"
  )

  # Let's extract the fold changes from `stats_df`
  log_fc_df <- stats_df_AA %>%
    # We only want to keep the `Gene` column as well
    dplyr::select("Metabolite", dplyr::contains("Basal")) %>%
  # Make this a longer format
    tidyr::pivot_longer(dplyr::contains("Basal"),
                      names_to = "contrast",
                      values_to = "logFoldChange"
    )


  # inner join these dataframes together
  plot_df <- log_fc_df %>%
    dplyr::inner_join(contrast_p_vals_df_AA,
                    by = c("Metabolite", "contrast"),
                    # This argument will add the given suffixes to the column names
                    # from the respective data frames, helping us keep track of which columns
                    # hold which types of values
                    suffix = c("_log_fc", "_p_val")
    )

  # Convert p value cutoff to negative log 10 scale
  p_val_cutoff <- -log10(0.05)

  # Absolute value cutoff for fold changes
  abs_fc_cutoff <- 0.5

  # use above cutoff to declare which are considered significant
  plot_df <- plot_df %>%
    dplyr::mutate(
      signif_label = dplyr::case_when(
        abs(logFoldChange) > abs_fc_cutoff & neg_log10_p_val > p_val_cutoff
        ~ "p-val and FC",
        abs(logFoldChange) > abs_fc_cutoff ~ "FC",
        neg_log10_p_val > p_val_cutoff ~ "p-val",
        TRUE ~ "NS"
      )
    )
  # convert contrast names to plot friendly
  plot_df <- mutate(plot_df, contrast = recode(contrast, "BasalD10_D1" = "Basal D10 v D1", "BasalD10_D7" = "Basal D10 v D7", "BasalD7_D1" = "Basal D7 v D1"))

  # now lets plot these as volcano plots for each contrast
  volcanoes_plot <- ggplot(
    plot_df,
    aes(
      x = logFoldChange, # Fold change as x value
      y = neg_log10_p_val, # -log10(p value) for the contrasts
      color = signif_label # Color code by significance cutoffs variable we made
    )
  ) +
    # Make a scatter plot with points that are 30% opaque using `alpha`
    geom_jitter(width = 0.06, alpha = 0.8, size = 3) +
    # Draw our `p_val_cutoff` for line here
    geom_hline(yintercept = p_val_cutoff, linetype = "dashed") +
    # Using our `abs_fc_cutoff` for our lines here
    geom_vline(xintercept = c(-abs_fc_cutoff, abs_fc_cutoff), linetype = "dashed") +
    # The default colors aren't great, we'll specify our own here
    scale_colour_manual(values = c("red", "darkgray", "blue", "black"), name = "Significance Label", breaks = c("FC", "NS", "p-val", "p-val and FC" )) +
    # Let's be more specific about what this p value is in our y axis label
    ylab("Contrast -log10(p value)") +
    # change y axis scale to match across plots
    ylim(0, 2.5) +
    # add title
    ggtitle(paste0("Significantly Different ", plot_id[[i]], " Metabolites")) +
    # This makes separate plots for each contrast!
    facet_wrap(~contrast) +
    # change facet wrap labels - https://www.datanovia.com/en/blog/how-to-change-ggplot-facet-labels/
    #facet_grid(~contrast, labeller(contrast = labs)) +
    # change x lab position
    theme(axis.text.x = element_text(angle = 45,vjust = 1, hjust=1))
    # Just for making it prettier!
    theme_classic()

  # Print out the plot!
  volcanoes_plot
  # cache plots in list
  volcanoPlot_list[[i]] <- volcanoes_plot
}

# arrange plots and write out
plotsArranged <- ggarrange(plotlist = volcanoPlot_list, nrow = 1, ncol = 1, common.legend = FALSE, legend = "bottom")
# export as PDFs
ggexport(plotsArranged, filename="/Users/danwilkinson/OneDrive - The University of Nottingham/Work_Folders/ResearchWork/PROJECT_ExtCollab/Zudin_ICU/Non_Normalised_ITUFeeding/FINAL_Data/AllMetabsFinal/EDA_figs/Volcano_Plots3_Rescaled.pdf")
```

## Plotting of Sig Output for Metabolites - Basal Comparisons and Correlation to Mass

```{r}
## Plots - creates individual plots per sig variable, create loop structure as before
# load in and merge with AA sig metabs
BasalSigMetabsSmallMol <- read_csv("/Users/mdzdw/Documents/WorkFolders/ResearchWork/PROJECT_ExtCollab/Zudin_ICU/Non_Normalised_ITUFeeding/FINAL_Data/SmallMolOnly/BasalTimeContrasts/BasalTimeSigSmallMolOnly.csv")
BasalSigMetabsAA <- read_csv("/Users/mdzdw/Documents/WorkFolders/ResearchWork/PROJECT_ExtCollab/Zudin_ICU/Non_Normalised_ITUFeeding/FINAL_Data/AAOnly/BasalTimeContrasts/BasalTimeSigAAOnly.csv")
# drop first col
BasalSigMetabsSmallMol <- BasalSigMetabsSmallMol[,-1]
BasalSigMetabsAA <- BasalSigMetabsAA[,-1]

# join in single dataframe
names(BasalSigMetabsAA)
join_byNames <- c("PtID","arm","Sample_ID","Day", "Sample_time","timePoint", "samplePoint", "site_id")
BasalTimeSigAllMetabs <- inner_join(BasalSigMetabsSmallMol, BasalSigMetabsAA, by = join_byNames) 

# make long dataframe from sigMetabs dataframe
sig_metabs <- names(BasalTimeSigAllMetabs)[9:ncol(BasalTimeSigAllMetabs)]
sigBasalMetabLonger <- as.data.frame(BasalTimeSigAllMetabs %>% pivot_longer(cols = all_of(sig_metabs), names_to = 'Metabs', values_to = 'NormAbundance'))

plot_list <- list() # to hold generated plots

for (i in 1:length(sig_metabs)) { # loop to create plots per metabolite
  metab <- sig_metabs[i]
  bxplt <- sigBasalMetabLonger %>% filter(Metabs == metab) %>% ggboxplot(x = "Day", y = "NormAbundance",
                 color = "Day", palette = "jco", x.text.angle = 45, ylab = "Log Abundance (AU)",
                 xlab = FALSE, legend = "bottom", title = paste0(metab), add = "jitter")
  # change font size and type
  bxplt <- bxplt + font("title", size = 12, face = "bold.italic") +
    font("ylab", size = 10) +
    font("x.text", size = 10) +
    font("y.text", size = 10)
  # cache plots in list
  plot_list[[i]] <- bxplt
} 

# arrange plots into 4 plots per page
plotsArranged <- ggarrange(plotlist = plot_list, nrow = 2, ncol = 2, common.legend = TRUE, legend = "bottom")
# export as PDFs
dir.create("/Users/mdzdw/Documents/WorkFolders/ResearchWork/PROJECT_ExtCollab/Zudin_ICU/Non_Normalised_ITUFeeding/FINAL_Data/AllMetabsFinal/EDA_figs")
ggexport(plotsArranged, filename="/Users/mdzdw/Documents/WorkFolders/ResearchWork/PROJECT_ExtCollab/Zudin_ICU/Non_Normalised_ITUFeeding/FINAL_Data/AllMetabsFinal/EDA_figs/SigMetabsChangeBasal_EDA_bxplots.pdf")
```

## Incorporate in physiological data - and plot correlations/lm

There is limited physiological data, therefore significant changes in Day 1 to Day 10 metabolites were plotted to change in muscle CSA, with baseline metabolites with SOFA scores

```{r}
# get sig metabs
metabs <- colnames(BasalTimeSigAllMetabs)[9:ncol(BasalTimeSigAllMetabs)]
# pivot sig metab data
allSigMetabsLong <- BasalTimeSigAllMetabs %>% pivot_longer(cols = `Acetone`:`Alanine_2`, names_to = 'metab', values_to = 'values')

# get day 1 & day 10 data for each metabolite
d1 <- allSigMetabsLong %>% filter(Day == 'd1')
BaselineMetabs <- allSigMetabsLong %>% filter(Day == 'd1') # for OrganFail Data Analyses
d10 <- allSigMetabsLong %>% filter(Day == 'd10')
# who has no day 10 data
no_d10 <- setdiff(d1$PtID, d10$PtID)
no_d1 <- setdiff(d10$PtID, d1$PtID)
outs <- c(no_d10, no_d1)
# drop these people from d1 dataset
d1 <- d1 %>% filter(!PtID %in% outs) %>% rename(values_1 = values)
d10 <- d10 %>% filter(!PtID %in% outs) %>% rename(values_10 = values)
# check same dimensions
dim(d1)
dim(d10)

# bind data back into single dataframe
allMetabsD1_D10 <- cbind(d1, d10$values_10) 
allMetabsD1_D10 <- rename(allMetabsD1_D10, values_10 = `d10$values_10`) # rename new cols

# create delta change variable for metabs
allMetabsD1_D10Diffs <- allMetabsD1_D10 %>% mutate(diffs = values_1 - values_10)

# load in phenotype data for patients
phenoData <- read_csv("/Users/mdzdw/Documents/WorkFolders/ResearchWork/PROJECT_ExtCollab/Zudin_ICU/Non_Normalised_ITUFeeding/IVC data for DW 250121.csv")

# match patient IDs to metab data for phenoData
d1_d10_pt_ids <- allMetabsD1_D10Diffs$PtID
phenoData <- phenoData %>% filter(UIN %in% d1_d10_pt_ids)

# create muscle CSA difference metric
phenoData <- phenoData %>% mutate(musc_diffs = `RFCSA D1 (%)` - `RFCSA D10 (%)`) %>% rename(PtID = UIN)
phenoData$PtID <- as.character(phenoData$PtID)
allMetabsD1_D10Diffs$PtID <- as.character(allMetabsD1_D10Diffs$PtID)

# join to create final dataframe for correl plots for phenoData
allD1_D10Diffs<- inner_join(allMetabsD1_D10Diffs, phenoData, by = 'PtID')
allD1_D10Diffs <- allD1_D10Diffs[,-c(13:22)]
head(allD1_D10Diffs)

# create loop to run through all metabs and save plot to list for CSA diffs
plot_listCorrel <- list() # to hold generated plots
metabName <- vector()
RSquared <- vector()
pVal <- vector()
metabs <- unique(allD1_D10Diffs$metab)

# loop for CSA diffs ~ metab diffs
for (i in 1:length(metabs)) { # loop to create plots per metabolite
  pltMetab <- metabs[i]
  plotDataFrame <- allD1_D10Diffs %>% filter(metab == pltMetab)
  res <- lm(musc_diffs ~ diffs, plotDataFrame) # run linear model
  sum <- summary(res) # extract summary model values
  plt <- ggplot(plotDataFrame, aes(diffs, musc_diffs)) + geom_point() + facet_wrap(~metab) + geom_smooth(method = "lm")
  plt <- plt + xlab("Day 1 - Day 10 Metabolite Abundance Change (AU)") + ylab("Day 1 - Day 10 Muscle CSA % Decline")
  plt <- plt + font("ylab", size = 10) +
    font("x.text", size = 10) +
    font("y.text", size = 10)
  # plt <- plt + ggtitle(paste0(pltMetab, "R squared"))
  # cache plots in list
  plot_listCorrel[[i]] <- plt
  metabName <- append(metabName, pltMetab)
  RSquared <- append(RSquared, sum$r.squared)
  pVal <- append(pVal, sum$coefficients[2,4])
} 

linearModelOut <- data.frame(Metabolite = metabName, 
                             RSquared = RSquared,
                             p_value = pVal) # linear model results table

# write out linear model results
write.csv(linearModelOut, "/Users/mdzdw/Documents/WorkFolders/ResearchWork/PROJECT_ExtCollab/Zudin_ICU/Non_Normalised_ITUFeeding/FINAL_Data/AllMetabsFinal/sigMetabLinearRegressionOutput.csv")
# arrange plots into 4 plots per page
plotsArranged <- ggarrange(plotlist = plot_listCorrel, nrow = 2, ncol = 1, common.legend = TRUE, legend = "bottom")
# export as PDFs
ggexport(plotsArranged, filename="/Users/mdzdw/Documents/WorkFolders/ResearchWork/PROJECT_ExtCollab/Zudin_ICU/Non_Normalised_ITUFeeding/FINAL_Data/AllMetabsFinal/EDA_figs/SigMetabsCorrelMassPlots.pdf")


### sofa scores with baseline metabs data ###

# load in organ failure score and link to baseline metabs
organFail <- read_csv("/Users/mdzdw/Documents/WorkFolders/ResearchWork/PROJECT_ExtCollab/Zudin_ICU/Non_Normalised_ITUFeeding/OrganFailureScoreNoValRemoved.csv")
organFail <- organFail[,-2] %>% rename(PtID = UIN)
# join to create final dataframe for correl plots for organ failure data
allBaselineOrganFail<- inner_join(BaselineMetabs, organFail, by = 'PtID')

# create loop to run through all metabs and save plot to list for SOFA scores
plot_listCorrelSofa <- list() # to hold generated plots
metabNameSofa <- vector()
RSquaredSofa <- vector()
pValSofa <- vector()
metabsSofa <- unique(allBaselineOrganFail$metab)

# loop for SOFA ~ metab baseline
for (i in 1:length(metabsSofa)) { # loop to create plots per metabolite
  pltMetab <- metabsSofa[i]
  plotDataFrame <- allBaselineOrganFail %>% filter(metab == pltMetab)
  res <- lm(`SOFA score` ~ values, plotDataFrame) # run linear model
  sum <- summary(res$coefficients) # extract summary model values
  plt <- ggplot(plotDataFrame, aes(values, `SOFA score`)) + geom_point() + facet_wrap(~metab) + geom_smooth(method = "lm")
  plt <- plt + xlab("Day 1 Metabolite Abundance Change (AU)") + ylab("SOFA Score")
  plt <- plt + font("ylab", size = 10) +
    font("x.text", size = 10) +
    font("y.text", size = 10)
  # plt <- plt + ggtitle(paste0(pltMetab, "R squared"))
  # cache plots in list
  plot_listCorrelSofa[[i]] <- plt
  metabNameSofa <- append(metabNameSofa, pltMetab)
  RSquaredSofa <- append(RSquaredSofa, sum$r.squared)
  pValSofa <- append(pValSofa, sum$coefficients[2,4])
} 

linearModelOutSofa <- data.frame(Metabolite = metabNameSofa, 
                             RSquared = RSquaredSofa,
                             p_value = pValSofa) # linear model results table

# write out linear model results
write.csv(linearModelOutSofa, "/Users/mdzdw/Documents/WorkFolders/ResearchWork/PROJECT_ExtCollab/Zudin_ICU/Non_Normalised_ITUFeeding/FINAL_Data/AllMetabsFinal/sigMetabLinearRegressionOutputSofa.csv")
# arrange plots into 4 plots per page
plotsArranged <- ggarrange(plotlist = plot_listCorrelSofa, nrow = 2, ncol = 1, common.legend = TRUE, legend = "bottom")
# export as PDFs
ggexport(plotsArranged, filename="/Users/mdzdw/Documents/WorkFolders/ResearchWork/PROJECT_ExtCollab/Zudin_ICU/Non_Normalised_ITUFeeding/FINAL_Data/AllMetabsFinal/EDA_figs/SigMetabsCorrelSofaPlots.pdf")


# load in csv files for table construction
## Sig Metabolite Linear Regression Output Table ##
linearModelOut <- read_csv("/Users/mdzdw/Documents/WorkFolders/ResearchWork/PROJECT_ExtCollab/Zudin_ICU/Non_Normalised_ITUFeeding/FINAL_Data/AllMetabsFinal/sigMetabLinearRegressionOutput_2.csv")
# remove column 1
linearModelOut <- linearModelOut[,-1]
# use SjPlot to export as HTML
tab_df(linearModelOut, title = "Linear Regression Analyses of Changing Metabolite Abundance Over 10 Days Against Muscle CSA Change", digits = 4, col.header = c("Metabolite", "R Squared", "p value"), file = "/Users/mdzdw/Documents/WorkFolders/ResearchWork/PROJECT_ExtCollab/Zudin_ICU/Non_Normalised_ITUFeeding/FINAL_Data/AllMetabsFinal/sigMetabMuscleDiff_lm")

# load in csv files for table construction
## Sig Metabolite Linear Regression SOFA Output Table ##
linearModelOutSofa <- read_csv("/Users/mdzdw/Documents/WorkFolders/ResearchWork/PROJECT_ExtCollab/Zudin_ICU/Non_Normalised_ITUFeeding/FINAL_Data/AllMetabsFinal/sigMetabLinearRegressionOutputSofa_2.csv")
# remove column 1
linearModelOutSofa <- linearModelOutSofa[,-1]
# Use SjPlot Here to export
tab_df(linearModelOutSofa, title = "Linear Regression Analyses of Baseline Metabolite Abundance Against SOFA Organ Failure Score", digits = 4,col.header = c("Metabolite", "R Squared", "p value"), file = "/Users/mdzdw/Documents/WorkFolders/ResearchWork/PROJECT_ExtCollab/Zudin_ICU/Non_Normalised_ITUFeeding/FINAL_Data/AllMetabsFinal/BaselineMetabSOFA_lm_2")

# load in csv files for table construction
## Sig Metab Summary Tables
linearModelOutSofa <- read_csv("/Users/mdzdw/Documents/WorkFolders/ResearchWork/PROJECT_ExtCollab/Zudin_ICU/Non_Normalised_ITUFeeding/FINAL_Data/AllMetabsFinal/sigMetabLinearRegressionOutputSofa_2.csv")
# remove column 1
linearModelOutSofa <- linearModelOutSofa[,-1]
# Use SjPlot Here to export
tab_df(linearModelOutSofa, title = "Linear Regression Analyses of Baseline Metabolite Abundance Against SOFA Organ Failure Score", digits = 4,col.header = c("Metabolite", "R Squared", "p value"), file = "/Users/mdzdw/Documents/WorkFolders/ResearchWork/PROJECT_ExtCollab/Zudin_ICU/Non_Normalised_ITUFeeding/FINAL_Data/AllMetabsFinal/BaselineMetabSOFA_lm_2")

# load in csv files for table construction
## Sig Metab Summary Tables
sigMetabOutput <- read_csv("/Users/mdzdw/Documents/WorkFolders/ResearchWork/PROJECT_ExtCollab/Zudin_ICU/Non_Normalised_ITUFeeding/FINAL_Data/AllMetabsFinal/sigMetabOutput_2.csv")
sigMetabOutput <- sigMetabOutput[,-c(7, 3)]
# Use SjPlot Here to export
tab_df(sigMetabOutput, title = "Significant Metabolites Changing in Abundance from D1 to D10", digits = 4,col.header = c("Metabolite", "log Fold Change Abundance", "t Statistic", "p value", "FDR Corrected p value", "HMDB Number", "Metabolic Pathway"), file = "/Users/mdzdw/Documents/WorkFolders/ResearchWork/PROJECT_ExtCollab/Zudin_ICU/Non_Normalised_ITUFeeding/FINAL_Data/AllMetabsFinal/SignificantMetabsAllTable")

#### END ####
```