---
title: "Non_NormlaisedNMR_ITU_Study - Initial Data Tidying and Visualisation"
author: "DJW and IJG"
date: "07/05/2020"
output: word_document
editor_options: 
  chunk_output_type: console
---

load libraries and packages
```{r}
library(tidyverse)
library(ggplot2)
library(ggfortify)
library(ggpubr)
library(pheatmap)
library(limma)
library(splines)
```

## Initial Data Tidying and Manipulation

load in non-normalised metabolite data & normalise by scaling and centering.
```{r}
setwd("/Users/mdzdw/Documents/WorkFolders/ResearchWork/PROJECT_ExtCollab/Zudin_ICU/Non_Normalised_ITUFeeding/")
# get data - change accordingly
#moleculeData <- read_csv("Plasma_SmallMolOnly_ITU_project.csv") # small mol
moleculeData <- read_csv("Plasma_Lipids_ITU_project.csv") # Lipids
#moleculeData <- read_csv("Plasma_AAOnly_ITU_project.csv") # AA

# quick rename
all_metab <- moleculeData
# some metabolite abundances are negative - remove these by converting to zeros
all_metab[all_metab < 0] <- 0
# log transform data + 1 to avoid inf or NaN's
all_metab[,2:ncol(all_metab)] <- log(all_metab[, 2:ncol(all_metab)] + 1) 
# check for NA's
sum(is.na(all_metab))
# scale and center
all_metab[,2:ncol(all_metab)] <- scale(all_metab[,2:ncol(all_metab)], scale = TRUE, center = TRUE) 
# check
sum(is.na(all_metab))
```

The molecule data are now scaled & centered. Next we tidy the data to standard long format for exploratory plots.

```{r}
# create long format data frame for plotting
namesCol <- names(all_metab[2:ncol(all_metab)])
all_metab_long <- all_metab %>% pivot_longer(cols=all_of(namesCol), names_to = "Metabs", values_to = "Peak_Area")
```

The study involved blood sampling from participants under 2 feeding regimes at various centers. Blood was sampled twice a day (9am & 1pm) on days 1, 7 & 10. Below we add information on time of sampling, day of sampling and center.

```{r}
# add in time, day and site information
sample_data <- read_csv('/Users/mdzdw/Documents/WorkFolders/ResearchWork/PROJECT_ExtCollab/Zudin_ICU/Plasma trial samples randomisation.csv',col_names=F)

# get a vector of the day of measurement
days <- as.character(sample_data[1,2:13]) 
# get a vector of the time of measurement
# format X_Y where X is time and Y is whether the sample was taken before (Y=1) or after the bolus of feed (Y=2)
other_value <- sample_data[2,2:13] 
# combine these and use as headers
new_col_names <- paste(as.character(days), as.character(other_value), sep=':')
new_col_names <- c('PtID', new_col_names)
# add new col names
colnames(sample_data) <- new_col_names
# drop redundant information from first 3 rows
sample_data <- sample_data %>% slice(3:n())

# pivot data
sample_data_long <- sample_data %>% pivot_longer(cols = `d1:9_1`:`d10:1_2`, names_to = c('Day', 'Sample_time'), names_sep=':', values_to = 'code')
# split out useful info in pt id, first two digits are for sites
# from Zudin - Patient id yes links back- the first two numbers to site, and the next three to the patient (email 28/2/2020)
sample_data_long$site_id <- str_sub(sample_data_long$PtID, 1, 2)
colnames(sample_data_long)[4] <- "Sample_ID"
sample_data_long <- na.omit(sample_data_long) # remove times for patients when samples weren't collected
```

Some participants have missing samples e.g. no plasma left and some IDs don't match to randomisation. We'll remove these.

```{r}
indexMissing <- setdiff(sample_data_long$Sample_ID, all_metab_long$Sample_ID)
# so we have some mismatch for reasons above

# filter out missing ids by keeping sample ids NOT IN(!) indexMissing
sample_data_long <- filter(sample_data_long, !Sample_ID %in% indexMissing)

# no differences? yes (i.e. 0) is good
setdiff(sample_data_long$Sample_ID, all_metab_long$Sample_ID)
# # should now be good to join
```

Merge site, time and day data with metabolite data.

```{r}
all_metab_long_Full <- full_join(all_metab_long, sample_data_long, by = "Sample_ID")
# check for NA
sum(is.na(all_metab_long_Full))
# where are these NA's?
missings <- apply(is.na(all_metab_long_Full), 2, which)
# there are 88 Na's in the datastet across cols PtID, Day, Sample_time, site_id
# remove these rows
all_metab_long_Full <- all_metab_long_Full[-missings$PtID,]
```

Add in information for which participant was in which arm of the study (continuous or bolus feeding).

```{r}
# treatment allocation info
alloc <- read_csv('/Users/mdzdw/Documents/WorkFolders/ResearchWork/PROJECT_ExtCollab/Zudin_ICU/Non_Normalised_ITUFeeding/treatment_allocation.csv', col_types = 'cc') 
# merge with all data
all_data <- inner_join(alloc, all_metab_long_Full, by = 'PtID') # inner_join only keeps those matching PtIDs
# data tidying complete - write out file change based on input file
#write.csv(all_data, "/Users/mdzdw/Documents/WorkFolders/ResearchWork/PROJECT_ExtCollab/Zudin_ICU/Non_Normalised_ITUFeeding/FINAL_Data/SmallMolOnly/allDataTidy_SmallMolNMR.csv") # change file name and address as appropriate.
write.csv(all_data, "/Users/mdzdw/Documents/WorkFolders/ResearchWork/PROJECT_ExtCollab/Zudin_ICU/Non_Normalised_ITUFeeding/FINAL_Data/LipidOnly/allDataTidy_LipidsNMR.csv") 
#write.csv(all_data, "/Users/mdzdw/Documents/WorkFolders/ResearchWork/PROJECT_ExtCollab/Zudin_ICU/Non_Normalised_ITUFeeding/FINAL_Data/AAOnly/allDataTidy_AANMR.csv") 

## END OF DATA TIDYING
```

